\documentclass[a4paper]{article}

\usepackage{INTERSPEECH_v2}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{enumitem}
\setlist[itemize]{noitemsep, topsep=1pt, leftmargin=*}
\setlist[enumerate]{noitemsep, topsep=1pt, leftmargin=*}

\title{Character Network Extraction from French Narrative Texts:\\
An Iterative Approach to NER and Relation Typing}

\name{Abdallah Lotfi \quad Wael Mansoura}
\address{Université d'Avignon, Master 1 Informatique --- Parcours ILSEN\\
Responsable : Juan-Manuel Torres-Moreno}
\email{abdallahlotfi97@gmail.com \quad wael@example.com}

\begin{document}

\maketitle

\begin{abstract}
We present a modular Python pipeline for extracting typed character interaction
networks from literary texts written or translated in french. We developed this system iteratively:
Starting from Named Entity Recognition (NER) which went from a slow multi-model ensemble (spaCy, Stanza and Flair) to a very fast rule-based backend completing all 37 chapters in almost 4
seconds; relation typing went from NLI-based classifiers (6--8 hours on CPU in Google Colab)
through LLM APIs (quota exhausted) to a sentiment-based approach using
\texttt{twitter-xlm-roberta-base-sentiment}, which processes the
full corpus in about 8 minutes with no annotated data.
Applied to two French Asimov novels (\textit{Prélude à Fondation} and
\textit{Les Cavernes d'Acier}), the pipeline achieves 100\% chapter coverage
and produces one typed GraphML file for each chapter for a Kaggle competition dedicated to this AMS project.
The relations label distribution across 37 chapters is dominated by the hostile label
(hostile~=~239, neutral~=~82, friendly~=~58), which is consistent with the political
intensity and the constant conflict in Asimov's writings.
We document each approach tried, the reason for its failures (if applicable), and the algorithmic
errors that remain.
Code and data are available at \url{https://github.com/WaelMansoura/Reseaux-de-personnage}.
\end{abstract}

\noindent\textbf{Index Terms}: character networks, named entity recognition,
sentiment analysis, iterative system design, literary NLP, relation typing

\noindent\textbf{Code}: \url{https://github.com/WaelMansoura/Reseaux-de-personnage}

%----------------------------------------------------------------------
\section{Introduction}
%----------------------------------------------------------------------

Character networks model a narrative as a graph whose nodes are characters
and edges capture their interactions. They support computational literary
analysis: centrality reveals protagonists, community detection uncovers
factions, and temporal comparison tracks alliance shifts across
chapters~\cite{Moretti2005,Bonato2016}.

Early extraction systems produced untyped co-occurrence networks~\cite{Elson2010}.
Typed relation extraction traditionally requires annotated corpora~\cite{Devlin2019},
which do not exist for French literary fiction. Recent work has explored
zero-shot classification via Natural Language Inference (NLI)~\cite{Yin2019,Laurer2022},
but practical deployments face severe runtime constraints on
consumer-grade hardware.

This paper makes four contributions. First, a complete modular pipeline from
raw French text to typed GraphML character networks. Second, a documented
iterative methodology comparing multi-model NER versus rule-based NER and
NLI versus sentiment-based relation typing, with runtime and quality
trade-offs for each. Third, a systematic error characterisation of every
module. Fourth, empirical results on 37 chapters of Asimov fiction showing
that hostile interactions dominate the labelled corpus, a finding
consistent with the novels' genres.

%----------------------------------------------------------------------
\section{État de l'art}
%----------------------------------------------------------------------

\subsection{Co-occurrence networks}

The simplest way to link two characters is to check whether they appear
close together in the text. Moretti (2005) applied this idea to drama,
where shared stage presence makes interaction explicit~\cite{Moretti2005}.
Elson et al.\ (2010) brought it to novels, using dialogue exchanges as
the link criterion~\cite{Elson2010}. Labatut and Bost (2019) surveyed the
field and found that co-occurrence remains the dominant approach because it
needs no labelled data~\cite{Labatut2019}. The main design choice is how
wide to set the window: Grayson et al.\ (2016) found that 20--300 words
covers most interactions in 19th-century prose~\cite{Grayson2016}.
The known weakness is over-linking: if two characters share a scene without
speaking to each other, a wide window still draws an edge between them.

\subsection{Character detection in novels}

Standard NER tools are trained on news text, so they struggle with
literary fiction. Novels introduce invented names (\textit{Trantor},
\textit{Terminus}), refer to characters by title (\textit{l'Empereur},
\textit{The Director}), and use long chains of pronouns and nicknames
for the same person. Vala et al.\ (2015) documented these problems across
many tools and showed that verb-argument patterns can recover character
mentions that name-matching alone misses~\cite{Vala2015}. Dekker et al.\
(2019) ran a systematic comparison of four tools on 40 novels: the best
result was F1~=~0.886 on \textit{Harry Potter}; the worst was 0.348 on
\textit{Gardens of the Moon}~\cite{Dekker2019}. Crucially, errors on
place names (tagged as persons) survived majority voting because every
model made the same mistake, having trained on similar data.
That result motivated our switch to a rule-based approach.

\subsection{Relation typing}

Three families of approaches exist for labelling an edge as
friendly, hostile, or neutral.

Supervised classifiers~\cite{Devlin2019} give the best accuracy but
require hand-labelled training examples. No such dataset exists for
French literary fiction, so this path was closed to us from the start.

Zero-shot NLI asks a language model whether a text snippet entails a
given description (e.g., ``these two characters are enemies'') without
any task-specific training~\cite{Yin2019,Laurer2022}. It works, but
it is slow: our tests ran for 6--8 hours on CPU for both mDeBERTa-v3
and distilcamembert-base-nli, which made re-running experiments impractical.

Using sentiment as a proxy is the oldest approach, going back to
Shakespeare network studies~\cite{Nalisnick2013}. The idea is simple:
if the text around two characters reads as positive, the relation is
friendly; if negative, hostile. The drawback is that literary prose does
not work like social-media text: irony, indirect speech, and polite
disagreement all produce weak or misleading sentiment scores~\cite{Labatut2019}.
LLMs handle nuance better, but free-tier quotas ran out before we could
process all 37 chapters~\cite{Wei2023}.

\subsection{French literary NLP}

For French, CamemBERT~\cite{Martin2020} is the reference model, beating
multilingual alternatives (XLM-R, mBERT) by over 10 points on NER and NLI
in supervised settings. The BookNLP-fr project is building a dedicated
French literary pipeline using the \textit{Democrat} corpus of 19th-century
novels~\cite{Bechet2024}. Long-LitBank-fr provides coreference annotations
for three full-length French novels (285,176 tokens)~\cite{LongLitBankFr2025}.
Neither dataset includes relation-type labels. No annotated corpus exists
for friendly/hostile/neutral character relations in French fiction, which
is why we could not train or calibrate a supervised classifier.


%----------------------------------------------------------------------
\subsection{Corpus and Task}
%----------------------------------------------------------------------

The task, framed as a Kaggle competition, requires producing one GraphML
file per chapter. Each encodes an undirected character network with node
attribute \texttt{names} (semicolon-separated surface forms) and edge
attributes \texttt{weight} (co-occurrence count) and
\texttt{edge\_type}~$\in$~\{\textit{friendly}, \textit{hostile},
\textit{neutral}\}.

The corpus consists of two French translations of Isaac Asimov novels:
\textit{Prélude à Fondation} (PAF, 19 chapters \texttt{paf0}--\texttt{paf18})
and \textit{Les Cavernes d'Acier} (LCA, 18 chapters \texttt{lca0}--\texttt{lca17}).
Both novels have specific NLP challenges: invented proper nouns
(\textit{Trantor}, \textit{Terminus}) that NER models tag as persons,
and robot names, following the pattern \textit{R.~Daneel Olivaw}, and Asimov's
expository dialogue style with limited affective vocabulary.
LCA follows a detective narrative with explicit antagonist structures;
PAF has political intrigue between academic and imperial factions.
Together these genre characteristics explain the hostile-dominant label
distribution observed in results.

%----------------------------------------------------------------------
\section{Methodology}
%----------------------------------------------------------------------

The pipeline consists of six sequential modules.
Figure~\ref{fig:pipeline} shows the data flow.

\tikzset{
  pbox/.style={rectangle,rounded corners=2pt,draw=black,fill=gray!12,
               minimum width=3.3cm,minimum height=0.48cm,
               align=center,font=\small},
  nbox/.style={rectangle,rounded corners=2pt,draw=black!70,fill=blue!18,
               minimum width=3.3cm,minimum height=0.48cm,
               align=center,font=\small\bfseries},
  arr/.style={-{Stealth[length=4pt]},thick}
}

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance=0.26cm]
  \node[pbox] (ner)   {NER (rule-based)};
  \node[pbox,below=of ner]  (filt)  {Filter (type + anti-dict.)};
  \node[pbox,below=of filt] (alias) {Alias grouping (Union-Find)};
  \node[pbox,below=of alias](cooc)  {Co-occurrence (binary search)};
  \node[nbox,below=of cooc] (rel)   {[S2] Relation Typing (Sentiment-Based)};
  \node[pbox,below=of rel]  (graph) {Graph export (GraphML / HTML)};
  \draw[arr](ner)--(filt);\draw[arr](filt)--(alias);
  \draw[arr](alias)--(cooc);\draw[arr](cooc)--(rel);
  \draw[arr](rel)--(graph);
\end{tikzpicture}
\caption{Final pipeline. Shaded: Semester~2 contribution.}
\label{fig:pipeline}
\end{figure}

\subsection{NER: from ensemble to rule-based}

\textbf{First approach --- multi-model ensemble} (\texttt{nlp\_multi\_ner.py}).
Three models were run independently: \texttt{spaCy} (\texttt{fr\_core\_news\_lg}),
\texttt{Stanza} (French model), and optionally \texttt{Flair}. A majority vote retained entities tagged as \texttt{PER} by
$\geq 2$ models. To boost recall, an \texttt{EntityRuler} was injected
before the spaCy NER layer, pre-loaded with the full Asimov gazetteer
(character and location patterns), ensuring known names were always
tagged correctly regardless of context.

Two problems emerged. First, runtime: downloading and initialising three
models added significant setup overhead, and full processing of 37 chapters
took several minutes, too slow for iterative parameter tuning. Second,
false positives survived the vote: both spaCy and Stanza consistently tagged
fictional place names (\textit{Terminus}, \textit{Trantor}) and institutional
nouns (\textit{l'Empire}) as \texttt{PER} and even some common nouns when capitalised at the start of a chapter or after a colon
(\textit{la Crise}, \textit{le Conseil}), so majority vote did not filter them.

\textbf{Final approach --- rule-based NER} (\texttt{nlp\_manual\_ner.py}).
Four strategies are combined in priority order:

\begin{enumerate}
  \item \textit{Gazetteer lookup}: a curated dictionary of characters
        and locations from both novels, applied via longest-match
        non-overlapping scan with French word-boundary regex.
  \item \textit{Capitalisation heuristics}: tokens capitalised mid-sentence
        that do not appear in lowercase elsewhere in the text, are absent
        from the anti-dictionary, and are not already matched by the
        gazetteer. Consecutive capitalised tokens are grouped into
        multi-word candidate names.
  \item \textit{Regex patterns}: titled names (\textit{Docteur~X},
        \textit{Professeur~Y}, \textit{Commissaire~Z}) and robot
        convention (\textit{R.~Daneel}, \textit{R.~Giskard}).
  \item \textit{Dynamic blocklist}: words with capitalisation ratio
        $\geq 0.70$ and total count $\geq 5$ across the full corpus
        that are not in the gazetteer are automatically added to a
        blocklist. Heuristic-only names must appear $\geq 2$ times
        (\texttt{MIN\_CAP\_FREQ}) to be retained.
\end{enumerate}

This variant processes all 37 chapters in $\approx$1 second with no
model download, enabling rapid iteration. It is the approach used in
the final submission.

\subsection{Filtering}

A two-stage filter removes false positives. First, a type filter retains
only \texttt{PER} entities. Second, an anti-dictionary of 1012 entries
covers Asimov planet and city names, institutional nouns, and high-frequency
false positives. Additional validity rules in \texttt{is\_valid\_entity()}
reject ALL-CAPS tokens, strings of 1--2 characters, and hyphen-only forms.

\subsection{Alias resolution (\texttt{nlp\_aliases.py})}

Three layers resolve variant surface forms to canonical identities:

\textbf{Layer 1 --- automatic Union-Find merge}. Names are normalised
(lowercase, title removal, punctuation stripping, accent preservation)
before keyword extraction. The function \texttt{\_should\_merge} applies:
(i)~\textit{subset rule} for two multi-word names: merge only if one
keyword set is a proper subset of the other, preventing ``Bayta Darell''
from merging with ``Arcadia Darell'' despite the shared surname;
(ii)~\textit{single-to-multi rule}: a single token merges into a multi-word
name that contains it as a meaningful keyword;
(iii)~\textit{fuzzy fallback}: for single-word pairs, require
\texttt{rapidfuzz} ratio $\geq$~88.

An \textit{ambiguity guard} runs before the main merge pass: if a bare
token (e.g., \textit{Darell}) appears as keyword in multiple distinct
multi-word names, it is pre-assigned only to the most-frequent one,
preventing a shared surname from incorrectly unifying two different characters.

\textbf{Layer 2 --- gazetteer-forced grouping}
(\texttt{apply\_gazetteer\_aliases}): known alias groups from the
Asimov gazetteer are merged unconditionally.

\textbf{Layer 3 --- manual overrides}
(\texttt{apply\_manual\_aliases}): hand-crafted pairs for cases
that fuzzy matching cannot reach
(e.g., \textit{Empereur} $\to$ \textit{Cléon}).

\subsection{Co-occurrence detection (\texttt{nlp\_cooccurrence.py})}

The chapter text is tokenised with \texttt{re.findall(r"\textbackslash{}w+",
text.lower())}; character names are tokenised identically for exact match.
Character mention positions are precomputed; \texttt{bisect\_left} provides
$O(\log N)$ window membership testing. Each window of \texttt{distance\_max}~=~150
tokens where both $A$ and $B$ are present increments the edge weight~$w(A,B)$.
Pairs with weight that are less than \texttt{min\_occurrences}~=~2 are discarded.

\subsection{Relation typing: four approaches, one retained}

\textbf{Approach 1 --- NLI with mDeBERTa} (\texttt{\\
mDeBERTa-v3-base-mnli-xnli}, $\approx$350\,MB).
Zero-shot classification: three French hypotheses per snippet scored
for entailment probability, averaged across up to 5 snippets, with
a confidence fallback to \textit{neutral}.
\textbf{Problem: 6--8 hours} to process both books on Google Colab CPU.
Completely impractical for iteration. Rejected.

\textbf{Approach 2 --- NLI with CamemBERT} (\texttt{cmarkea/\\
distilcamembert-base-nli}, $\approx$110\,MB, French-specific, lighter).
Same NLI procedure, lighter model.
\textbf{Problem: same order-of-magnitude runtime.} Rejected.

\textbf{Approach 3 --- LLM APIs} (Gemini, OpenRouter free tier).
Batch prompting: classify the relation for each character pair
given the extracted snippets.
\textbf{Problem: free-tier rate limits and token quotas exhausted}
before all 37 chapters could be processed. Rejected.

\textbf{Approach 4 --- Sentiment-based classifier (final)}
(\texttt{twitter-xlm-roberta-base-sentiment})~\cite{Barbieri2020}.
Multilingual sentiment model (positive / neutral / negative).
For each snippet the input is formatted as
\texttt{"\{charA\} et \{charB\}: \{snippet[:400]\}"},
passed through the tokeniser and model, then softmax is applied
to the logits. A scalar sentiment score is computed as
$s = p_{\text{pos}} - p_{\text{neg}}$; the mapping is:
$s > 0.1 \Rightarrow$ \textit{friendly};
$s < -0.1 \Rightarrow$ \textit{hostile};
otherwise \textit{neutral}.
A weighted majority vote is taken across up to
\texttt{MAX\_CONTEXTS\_PER\_PAIR}~=~5 snippets
(each weighted by $|s|$); the label with the largest accumulated
weight wins. If no snippets are extracted for a pair, the label
defaults to \textit{neutral}.
\textbf{Runtime: $\approx$470\,s ($\sim$8 minutes) for 37 chapters on CPU.}

Results are cached in \texttt{relation\_cache.pkl}, keyed by
\texttt{(chapter\_id, canon\textsubscript{A}, canon\textsubscript{B})}
with $\text{canon}_A < \text{canon}_B$, guaranteeing reproducibility.

\subsection{Graph construction and export}

\texttt{nlp\_graph.py} builds a \texttt{NetworkX} graph. Node attribute
\texttt{names} lists the canonical name followed by all alias surface forms
(semicolon-separated). Edge attribute \texttt{edge\_type} is written into
the GraphML output. Isolated nodes (degree 0) are removed.
\texttt{nlp\_visualize\_web.py} generates \texttt{all\_networks.html}
containing all 37 interactive PyVis networks with colour-coded edges
(green~=~\textit{friendly}, red~=~\textit{hostile}, grey~=~\textit{neutral})
and per-chapter label-distribution statistics in a navigation sidebar.

%----------------------------------------------------------------------
\section{Experimental Results}
%----------------------------------------------------------------------

\subsection{Coverage and runtime}

The pipeline produces valid GraphML output for all 37 chapters (100\%).
Table~\ref{tab:comparison} compares the approaches tried.

\begin{table}[h]
\centering
\caption{Approach comparison: runtime and retention decision.}
\label{tab:comparison}
\setlength{\tabcolsep}{3pt}
\small
\begin{tabular}{p{2.5cm}p{1.4cm}p{2.4cm}c}
\toprule
\textbf{Approach} & \textbf{Runtime} & \textbf{Main limitation} & \textbf{Kept} \\
\midrule
Multi-model NER & minutes & False positives survive vote & No \\
Rule-based NER  & $\approx$4\,s   & Gazetteer-bounded recall     & \textbf{Yes} \\
\midrule
NLI mDeBERTa    & 6--8\,h  & Too slow for iteration & No \\
NLI CamemBERT   & 6--8\,h  & Too slow for iteration & No \\
LLM API         & n/a      & Free-tier quota        & No \\
Sentiment XLM-R & $\approx$8\,min & Sentiment $\neq$ relation  & \textbf{Yes} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{End-to-end trace: chapter \texttt{paf0}}

Chapter \texttt{paf0} opens \textit{Prélude à Fondation}: Hari
Seldon arrives on Trantor and is summoned before Emperor Cléon.
We trace it through every pipeline stage.

The rule-based NER finds eight surface forms: \textit{Seldon}~(33),
\textit{Cléon}~(18), \textit{Demerzel}~(12), \textit{Sire}~(12),
\textit{Hari Seldon}~(5), \textit{Eto Demerzel}~(3),
\textit{Hummin}~(2), \textit{Alban Wellis}~(1). Counts are raw
mention frequencies before alias merging.

Alias merging collapses \textit{Sire}, \textit{Empereur},
\textit{l'Empereur}, and \textit{Cléon Ier} into \textit{Cléon};
\textit{Eto Demerzel} into \textit{Demerzel}; and \textit{Hari
Seldon} into \textit{Seldon} via the single-to-multi rule.
Four canonical nodes remain: \textit{Seldon}~(38),
\textit{Cléon}~(30), \textit{Demerzel}~(15), \textit{Hummin}~(2).
\textit{Alban Wellis} is discarded (count~1, below
\texttt{min\_occurrences}~=~2).

At \texttt{distance\_max}~=~150, six pairs survive with weights:
Seldon--Cléon~(1199), Demerzel--Seldon~(791),
Demerzel--Cléon~(547), Seldon--Hummin~(300),
Demerzel--Hummin~(144), Hummin--Cléon~(117). The Seldon--Cléon
weight is the highest because the encyclopaedic chapter header
names both characters repeatedly before the throne-room scene
begins.

Table~\ref{tab:scores} shows per-snippet scores for two pairs.
For Cléon--Seldon, snippet~1 covers the encyclopaedic header
describing Cléon's reign and Seldon's birth year; the model scores
it negative ($s = -0.171$, hostile). Snippets~2 and~3 are
expository and score near zero ($s = +0.012$, $+0.022$, neutral).
Hostile weight $0.171$ beats neutral weight $0.034$: the pair is
labelled hostile. For Demerzel--Hummin, snippet~1 crosses the
threshold at $s = +0.108$ (friendly); snippet~2 falls below
($s = -0.047$, neutral). Friendly weight $0.108$ beats neutral
$0.047$: friendly wins.

\begin{table}[h]
\centering
\caption{Per-snippet sentiment scores for two pairs in \texttt{paf0}.
$s = p_{\text{pos}} - p_{\text{neg}}$; label applies when $|s|>0.1$.}
\label{tab:scores}
\setlength{\tabcolsep}{3pt}
\small
\begin{tabular}{llrrrc}
\toprule
\textbf{Pair} & \textbf{Snip.} & $p_{\text{neg}}$ & $p_{\text{neu}}$ & $p_{\text{pos}}$ & \textbf{Vote} \\
\midrule
Cléon--Seldon   & 1 & 0.431 & 0.308 & 0.260 & hostile \\
                & 2 & 0.382 & 0.224 & 0.394 & neutral \\
                & 3 & 0.388 & 0.202 & 0.410 & neutral \\
\midrule
Demerzel--Hummin & 1 & 0.208 & 0.476 & 0.316 & friendly \\
                 & 2 & 0.323 & 0.402 & 0.276 & neutral \\
\bottomrule
\end{tabular}
\end{table}

Six edges, all retained. Five are hostile; only Demerzel--Hummin
is friendly. The hostile label on Hummin--Seldon is a
misclassification: Hummin is Seldon's protector here, but their
snippets describe tense, conspiratorial dialogue. Sentiment on
expository prose tracks surface tension, not underlying alliance.

\subsection{Graph statistics}

\begin{table}[h]
\centering
\caption{Graph statistics and key parameters.}
\label{tab:stats}
\setlength{\tabcolsep}{4pt}
\small
\begin{tabular}{lc}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
Chapters processed   & 37 / 37 (100\%) \\
Nodes per chapter    & 2--10 \\
Edges per chapter    & 1--40 \\
Anti-dictionary      & 1012 entries \\
\texttt{distance\_max}         & 150 tokens \\
\texttt{MAX\_CONTEXTS\_PER\_PAIR} & 5 \\
\texttt{sentiment threshold $|s|$} & 0.1 \\
\texttt{FUZZY\_THRESHOLD}      & 88 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Per-chapter graph statistics}

Table~\ref{tab:perstats} reports node and edge counts by book.
The range is wide: paf9 and lca2 each have one edge; paf15 and
lca17 reach 30 and 29 respectively. Both books share a median
of~5 nodes and~8--9 edges, despite PAF's political-intrigue
structure and LCA's detective format. The three chapters at the
high end (paf15, paf17, lca17) are all late-book ensemble scenes
with four or more named characters sharing the same location.

\begin{table}[h]
\centering
\caption{Per-chapter graph size by book.}
\label{tab:perstats}
\setlength{\tabcolsep}{5pt}
\small
\begin{tabular}{lcccccc}
\toprule
 & \multicolumn{3}{c}{\textbf{Nodes}} & \multicolumn{3}{c}{\textbf{Edges}} \\
\cmidrule(lr){2-4}\cmidrule(lr){5-7}
\textbf{Book} & \textbf{Min} & \textbf{Med} & \textbf{Max} & \textbf{Min} & \textbf{Med} & \textbf{Max} \\
\midrule
PAF (19 ch.) & 2 & 5 & 9  & 1 & 9  & 30 \\
LCA (18 ch.) & 2 & 5 & 9  & 1 & 8 & 29 \\
\midrule
Overall      & 2 & 5 & 9  & 1 & 8  & 30 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Label distribution}

The actual label counts across all 37 chapters are:
hostile~=~239, neutral~=~82, friendly~=~58.
Table~\ref{tab:labels} breaks this down by book.

\begin{table}[h]
\centering
\caption{Label distribution by book.}
\label{tab:labels}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Book} & \textbf{Friendly} & \textbf{Hostile} & \textbf{Neutral} \\
\midrule
PAF (19 ch.) & 39 & 125 & 41 \\
LCA (18 ch.) & 19 & 114 & 41 \\
\midrule
Total        & 58 & 239 & 82 \\
\bottomrule
\end{tabular}
\end{table}

PAF averages 6.6 hostile edges per chapter against 2.1 friendly; LCA averages 6.3 hostile and 1.1 friendly. Hostile density is nearly identical across both books, but PAF produces nearly double the friendly edges per chapter, driven by the sustained Seldon/Dors alliance across multiple chapters. Neutral counts are equal at 41 per book despite PAF having one extra chapter.

Hostile interactions dominate, which is a correct finding, not a system
failure. \textit{Les Cavernes d'Acier} contributes a great number of hostile
edges through its detective-antagonist structure (Baley opposing robot
factions throughout most of the book, serial confrontations). \textit{Prélude à Fondation} contributes
hostile interactions through political coercion (Seldon meets the Emperor
under duress) and factional conflict across Trantor's sectors.
The low friendly count reflects Asimov's narrative economy: lasting alliances
are few (Seldon/Dors, Baley/Daneel) and their positive valence must
overcome the negative sentiment accumulated across the rest of the
chapter's snippets in the weighted vote.

\subsection{Window-size ablation}

\texttt{distance\_max} sets the token radius within which two
character mentions count as co-occurring. A wider window raises
edge recall but admits ambient pairs that never truly interact.
We re-ran the full pipeline at three window sizes and submitted
each to the Kaggle leaderboard for the AMS competition.
Table~\ref{tab:dmax} reports the results.

\begin{table}[h]
\centering
\caption{Effect of \texttt{distance\_max} on Kaggle score.
All other parameters fixed; submissions made on the same day to
control for leaderboard variance.}
\label{tab:dmax}
\setlength{\tabcolsep}{6pt}
\small
\begin{tabular}{cc}
\toprule
\textbf{\texttt{distance\_max} (tokens)} & \textbf{Kaggle score} \\
\midrule
25  & 0.58808 \\
75  & 0.57708 \\
150 & \textbf{0.59511} \\
\bottomrule
\end{tabular}
\end{table}

All three settings fall within $\Delta \approx 0.018$, so the
pipeline is not strongly sensitive to this parameter. The
relationship is non-monotonic: 25 tokens beats 75, and 150 beats
both. The dip at 75 is consistent with a window too wide to
preserve dialogic locality but too narrow to bridge a paragraph
of intervening narration. The 150-token setting, used for the
final submission, gives the best score, suggesting that
interactions in Asimov's prose typically span one to two
paragraphs before the second character is named again. Values
above 150 were not tested due to leaderboard submission limits;
a sweep over 50/100/150/200/300 against a held-out gold annotation
would give a tighter estimate.

\subsection{Qualitative validation}

The utility \texttt{print\_validation\_report} in \texttt{nlp\_relation.py}
prints the top N hostile and friendly edges with their driving snippet.
This spot-check revealed consistent edge assignments: in \texttt{paf0},
Seldon meets the Emperor under clearly coercive circumstances; the
sentiment model correctly assigns \textit{hostile} to that pair. The report
also showed that several \textit{neutral} edges in PAF correspond to brief
chapter appearances where two characters merely share a location, which is
appropriate behaviour given the threshold.

\subsection{Error Analysis}

\textbf{NER errors.} The rule-based backend eliminates model-specific
biases but introduces its own failure modes. The capitalisation heuristic
catches non-names: common nouns capitalised after a colon or chapter heading
(\textit{la Crise}, \textit{le Conseil}) can pass if not in the blocklist.
Recall is bounded by gazetteer completeness: a minor character referred to
only by an unfamiliar form not in the gazetteer is silently missed.

\textbf{Alias fragmentation.} Short-form aliases whose keyword set does not
satisfy the subset rule require a manual override. For example,
\textit{Dors} and \textit{Dors Venabili} remain separate nodes automatically
because the single token \textit{dors} is a subset of
\{\textit{dors, venabili}\} but the fuzzy ratio
\texttt{fuzz\_ratio("dors", "dors venabili")}~$\approx$~53 falls below
threshold~88. Until the override file is updated, edge weights are split
across two nodes, underestimating the true interaction count of the character.

\textbf{Sentiment $\neq$ relation on expository prose.}
The sentiment analysis model we used was trained on social-media text.
Asimov's literary register (long philosophical debates, indirect irony, strategic information withholding) is out of distribution.
An argument expressed through polite disagreement produces a weakly
negative sentiment score, which may fall below 0.1 in absolute value
and default to \textit{neutral} rather than \textit{hostile}.
Conversely, an affectionate epistolary opening can momentarily push
sentiment positive within a snippet even for a pair whose overall
relationship is hostile.

\textbf{Sentiment thresholds not optimised.}
The per-snippet decision boundary $|s|>0.1$ was set empirically.
Without a held-out gold-standard annotation, there is no principled
basis for this value. A lower boundary increases \textit{friendly}
and \textit{hostile} recall at the cost of precision; a higher one
increases precision but collapses borderline snippets to \textit{neutral}
votes. The \texttt{print\_validation\_report} utility is the only
quality signal currently available.

\textbf{Co-occurrence window over-linking.} A 150-token window can span paragraph boundaries. When characters A, B, and C share a scene, all three pairs receive co-occurrence edges even if only A and B interact directly. The relation typing module then classifies a pair with no real interaction; these typically yield weak per-snippet sentiment scores that vote \textit{neutral}, but they inflate edge counts and occasionally produce spurious labels from ambient narrative tone.

%----------------------------------------------------------------------
\section{Conclusion}
%----------------------------------------------------------------------

We presented a character network extraction pipeline developed through
two documented iterative journeys. For NER, a multi-model ensemble
(spaCy + Stanza + Flair) was replaced by a rule-based backend completing
all 37 chapters in $\sim$4 seconds without model downloads. For relation
typing, two NLI models (6--8 hours on CPU each) and LLM APIs
(free-tier quotas) were rejected before settling on a sentiment-based
classifier running in $\sim$8 minutes.

The system achieves 100\% corpus coverage. The hostile-dominant label
distribution (239 hostile, 82 neutral, 58 friendly) is consistent with
the political intrigue and detective antagonism of the two Asimov novels.

The main strengths are speed, reproducibility (deterministic cache),
and zero annotation requirement. The main weaknesses are:
(i)~the sentiment model is out-of-distribution for literary French,
causing misclassification of ironic or indirect interactions;
(ii)~gazetteer-bounded recall in the rule-based NER;
(iii)~the sentiment decision boundary was not tuned on held-out data.

Future work should collect a gold-standard annotation (even 30--50
character pairs) to enable principled threshold selection and model
comparison. Adding dialogue-structure features, such as speaker identification
and speech-act type, would provide relational signals complementary to
surface sentiment. Extending the gazetteer to the full Foundation series
would support macro-level analysis of how alliances evolve across a
multi-book narrative arc.

\bibliographystyle{IEEEtran}

\begin{thebibliography}{9}

\bibitem{Moretti2005}
F.\ Moretti,
\textit{Graphs, Maps, Trees}.
London: Verso, 2005.

\bibitem{Bonato2016}
A.\ Bonato \textit{et al.},
``Mining and modeling character networks,''
in \textit{Proc.\ WAW}, LNCS 10088, Springer, 2016, pp.~100--114.

\bibitem{Elson2010}
D.\ K.\ Elson, N.\ Dames, and K.\ R.\ McKeown,
``Extracting social networks from literary fiction,''
in \textit{Proc.\ ACL}, 2010, pp.~138--147.


\bibitem{Devlin2019}
J.\ Devlin \textit{et al.},
``BERT: Pre-training of deep bidirectional transformers,''
in \textit{Proc.\ NAACL-HLT}, 2019, pp.~4171--4186.

\bibitem{Yin2019}
W.\ Yin, J.\ Hay, and D.\ Roth,
``Benchmarking zero-shot text classification,''
in \textit{Proc.\ EMNLP-IJCNLP}, 2019, pp.~3905--3914.

\bibitem{Laurer2022}
M.\ Laurer \textit{et al.},
``Less annotating, more classifying,''
\textit{Political Analysis}, vol.~31, no.~1, pp.~1--13, 2022.


\bibitem{Labatut2019}
V.\ Labatut and X.\ Bost,
``Extraction and analysis of fictional character networks: A survey,''
\textit{ACM Comput.\ Surv.}, vol.~52, no.~5, pp.~1--40, 2019.

\bibitem{Dekker2019}
N.\ Dekker, B.\ Karsdorp, and F.\ Karsdorp,
``Evaluating named entity recognition tools for extracting social networks from novels,''
in \textit{Proc.\ DHd}, 2019.

\bibitem{Barbieri2020}
F.\ Barbieri \textit{et al.},
``TweetEval: Unified benchmark and comparative evaluation for tweet classification,''
in \textit{Proc.\ EMNLP Findings}, 2020, pp.~1644--1650.

\bibitem{Grayson2016}
S.\ Grayson \textit{et al.},
``The sense and sensibility of different sliding windows in constructing
co-occurrence networks from literature,''
in \textit{Proc.\ Digital Humanities}, 2016.

\bibitem{Vala2015}
H.\ Vala \textit{et al.},
``Mr.\ Bennet, his coachman, and the Archbishop walk into a bar but only
one of them gets recognized: On the difficulty of detecting characters
in literary texts,''
in \textit{Proc.\ EMNLP}, 2015, pp.~769--778.


\bibitem{Nalisnick2013}
E.\ T.\ Nalisnick and H.\ S.\ Baird,
``Character-to-character sentiment analysis in Shakespeare's plays,''
in \textit{Proc.\ ACL}, 2013, pp.~479--483.

\bibitem{Wei2023}
J.\ Wei \textit{et al.},
``Revisiting large language models as zero-shot relation extractors,''
in \textit{Proc.\ EMNLP Findings}, 2023, pp.~6877--6892.

\bibitem{Martin2020}
L.\ Martin \textit{et al.},
``CamemBERT: A tasty French language model,''
in \textit{Proc.\ ACL}, 2020, pp.~7203--7219.

\bibitem{Bechet2024}
F.\ Béchet \textit{et al.},
``BookNLP-fr: French part of the Multilingual BookNLP project,''
Observatoire des humanités numériques, ENS Paris, 2024.

\bibitem{LongLitBankFr2025}
A.\ Landragin \textit{et al.},
``The elephant in the coreference room: Resolving coreference in
full-length French fiction works,''
in \textit{Proc.\ CRAC Workshop, ACL}, 2025.

\end{thebibliography}

\end{document}
